Comparing Power Optimizations in the Reorder Buffer

### 2011-06-08 paper/bg.tex
paper, rob latch update
Paper stuff. Put printouts behind m_print flag (enable with -d)

Upon execution completion, the output value is stored in the ROB with its 
corresponding entry and that entry is marked as valid or completed with a 
1 bit flag.  An entry is committed from the ROB to the architectural register
file (ARF) when that entry is the oldest (lowest PC) and its result has been 
marked as valid. 
\subsection{Circular ROB}
The circular ROB is one of the most basic implementations of the ROB.  Simply using
a circular buffer as the structure to hold each entry.  Since the ROB is basically
just memory, there are two special slots for to keep track of the head and tail
pointer to create the circular buffer.  As entries are written into the ROB, the
entry is written at the tail pointer and then the tail pointer is incremented.
Likewise when an entries is written to the ARF, the head pointer is incremented.
The circular buffer method is widely used since it inherently keeps track of the 
oldest instruction (which is at the head pointer), as well as which entry slots
are used or empty.   However it may not be very power efficient if there is a large
number of empty slots at any given point of the execution.    
\subsection{Dynamic ROB}
The dynamic ROB aims to reduce power by turning off portions of the buffer when they
are not needed, and turning them back on when they become necessary. By partitioning
the buffer into what is essentially smaller buffers, the complexity of each partition
is reduced. Further, the ROB will then be custom sized to need, thus reducing energy
wasted powering large stretches of empty entries. Due to the large percentage of energy
consumption stemming from leakage power, turning off partitions contribute significant
power savings over the base circular ROB.

@@

The distribution or clustering of the ROB unit also inherently induces a minimal performance overhead 
in particular cases.  By breaking the ROB into smaller pieces and distributing the instruction 
data across the unit, the placement of instructions becomes a performance factor.  Typically, a 
distributed ROB model features only a single write port per ROB cluster.  As a result only one operand 
can be forwarded at a time from the ROB.  In the rare case that an instruction requires two operands be 
forwarded from the same ROB cluster a delay of one cycle is imposed while the operands are forwarded 
sequentially.  It is reported, however, that this happens very rarely in practice and induces less than 1\% 
performance degradation \cite{rabaey}.
\subsection{Latch ROB}
In addition to large power requirement to maintain the entries of the ROB, another 
large power consumer is the read ports of the ROB.  The read port of the ROB is used
to forward the results of an entry to the dispatch unit (DU) to use as an operand.
All though forwarding from the ROB occurs about 6\%\cite{kucuk} of all instructions 
in an N way super scalar processor there are 2N ports (to accommodate FP
 instructions) which add unnecessary complexity and power consumption for a small 
percentage of instructions.  It has been proposed by eliminating the read ports 
in the ROB all together, we can reduce energy consumption\cite{kucuk}.  However
without read ports, a small window appears where the data cannot be accessed,
starting from when the result is placed in to the ROB until the ROB commits that 
entry to the ARF.  To help mitigate stalls that might happen due to this hole, 
we introduce retention latches.   Retention latches are small memory elements and 
we use them to cache the most recent results that have been written into the ROB.
This way instructions are still able to access the most recently produced results
and not stall.  However if the operand needed is not in the ROB, the stream will 
stall until a commit is made to the ARF, and that result value is forwarded to 
the dispatch unit immediately.

>>>>

Upon execution completion, the output value is stored in the ROB with its 
corresponding entry and that entry is marked as valid or completed with a 
1 bit flag.  An entry is committed from the ROB to the architectural register
file (ARF) when that entry is the oldest (lowest PC) and its result has been 
marked as valid. 
\subsection{Circular ROB}
The circular ROB is one of the most basic implementations of the ROB.  Simply using
a circular buffer as the structure to hold each entry.  Since the ROB is basically
just memory, there are two special slots for to keep track of the head and tail
pointer to create the circular buffer.  As entries are written into the ROB, the
entry is written at the tail pointer and then the tail pointer is incremented.
Likewise when an entries is written to the ARF, the head pointer is incremented.
The circular buffer method is widely used since it inherently keeps track of the 
oldest instruction (which is at the head pointer), as well as which entry slots
are used or empty.   However it may not be very power efficient if there is a large
number of empty slots at any given point of the execution.    
\subsection{Dynamic ROB}

The dynamic ROB aims to reduce power usage by turning off portions of the buffer when they are not needed, and turning them back on when they become necessary. By partitioning the buffer into what is essentially smaller buffers, the complexity of each partition is reduced. Further, the ROB will then be custom sized to need, thus reducing energy wasted powering large stretches of empty entries. Due to the large percentage of energy consumption stemming from leakage power, turning off partitions contribute significant power savings over the base circular ROB.

Downsizing the ROB is based on a periodic sampling of the buffer state. A number of cycles, {\it update\_period}, is defined, and every {\it update\_period} cycles the size of the ROB is updated. Every {\it sample\_period} cycles, the number of active entries is recorded, and the quantity {\it active\_size} is recorded as the average of these samples between each {\it update\_period}. The current maximum capacity of the reorder buffer is defined as the {\it current\_size}, and the size of a single ROB partition is defined as {\it partition\_size}. After each {\it update\_period} cycles, the ROB checks how full the buffer is, and if {\it current\_size} - {\it active\_size} > {\it partition\_size}, then a single partition is turned off.

Increasing the size of the ROB is dependent on how often the buffer fills up. A value, defined as the {\it overflow\_count} is incremented each cycle that instruction issue is stalled because the ROB is full. An {\it overflow\_threshold} is defined, and if during an {\it update\_period}, {\it overflow\_count} > {\it overflow\_threshold}, then a partition is turned on.

@@

The distribution or clustering of the ROB unit also inherently induces a minimal performance overhead 
in particular cases.  By breaking the ROB into smaller pieces and distributing the instruction 
data across the unit, the placement of instructions becomes a performance factor.  Typically, a 
distributed ROB model features only a single write port per ROB cluster.  As a result only one operand 
can be forwarded at a time from the ROB.  In the rare case that an instruction requires two operands be 
forwarded from the same ROB cluster a delay of one cycle is imposed while the operands are forwarded 
sequentially.  It is reported, however, that this happens very rarely in practice and induces less than 1\% 
performance degradation \cite{rabaey}.
\subsection{Latch ROB}
In addition to large power requirement to maintain the entries of the ROB, another 
large power consumer is the read ports of the ROB.  The read port of the ROB is used
to forward the results of an entry to the dispatch unit (DU) to use as an operand.
All though forwarding from the ROB occurs about 6\%\cite{kucuk} of all instructions 
in an N way super scalar processor there are 2N ports (to accommodate FP
 instructions) which add unnecessary complexity and power consumption for a small 
percentage of instructions.  It has been proposed by eliminating the read ports 
in the ROB all together, we can reduce energy consumption\cite{kucuk}.

However without read ports, a small window appears where the data cannot be accessed,
starting from when the result is placed in to the ROB until the ROB commits that 
entry to the ARF.  To help mitigate stalls that might happen due to this hole, 
we introduce retention latches.   Retention latches are small memory elements and 
we use them to cache the most recent results that have been written into the ROB.
This way instructions are still able to access the most recently produced results
and not stall.  However if the operand needed is not in the ROB, the stream will 
stall until a commit is made to the ARF, and that result value is forwarded to 
the dispatch unit immediately.

### 2011-06-08 paper/proj.tex
Paper stuff. Put printouts behind m_print flag (enable with -d)

Downsizing the ROB is based on a periodic sampling of the buffer state. A number of cycles, {\it update\_period}, is defined, and every {\it update\_period} cycles the size of the ROB is updated. Every {\it sample\_period} cycles, the number of active entries is recorded, and the quantity {\it active\_size} is recorded as the average of these samples between each {\it update\_period}. The current maximum capacity of the reorder buffer is defined as the {\it current\_size}, and the size of a single ROB partition is defined as {\it partition\_size}. After each {\it update\_period} cycles, the ROB checks how full the buffer is, and if {\it current\_size} - {\it active\_size} > {\it partition\_size}, then a single partition is turned off.

Increasing the size of the ROB is dependent on how often the buffer fills up. A value, defined as the {\it overflow\_count} is incremented each cycle that instruction issue is stalled because the ROB is full. An {\it overflow\_threshold} is defined, and if during an {\it update\_period}, {\it overflow\_count} > {\it overflow\_threshold}, then a partition is turned on.

@@

As a few minor changes to the behavioral function of the OOP ROB, the first of which is the lack of the ability to forward operands.  Thus in the simulated stage of reading from the instruction
stream, when we would normally check if the operand is in the ROB and forward.  We instead do the following, we still check to see if the operand is in the ROB, and the retention latches. 
If the entry is not in the ROB, or the latches, the operand would be available in the ARF or memory, in the converse case,  the entry in both, the ROB, and the latches, the operand is forwarded
from the latches and the counter is incremented .  In the case that the entry exists in the ROB, but not in the latches, a flag would be set in the ROB to forward the result when committing to the ARF. 
The last case, that is the entry exists in the latches and not in the ROB, means that the result has already been committed and the result could be forwarded from the ARF.  Next in our update entries 
(when the cycle counter expires -- execution has completed) we would add in the new entry to the retention latch.  Lastly when the ROB entires are valid, and ready to be committed to the ARF,
if the forwarding flag is set, the forwarding counter is incremented.
\subsection{Power Tabulation Implementation}
In order for our simulation to determine the power utilization, we do not try and equate each event with an associated cost in joules.  As technology changes, and new techniques are 
developed to decease circuit level power usages, our simulation will become less and less accurate.  Instead we chose to simply tabulate each event that is associated with 
power cost, such as bit flips from 0 to 1, register comparator usage, and bit lines driven.  To help measure leakage power and overall states of the ROB, we also keep track of the 
time (in cycles) each bit stays in the high state, or the low state.

>>>>

@@

As a few minor changes to the behavioral function of the OOP ROB, the first of which is the lack of the ability to forward operands.  Thus in the simulated stage of reading from the instruction stream, when we would normally check if the operand is in the ROB and forward.  We instead do the following, we still check to see if the operand is in the ROB, and the retention latches. If the entry is not in the ROB, or the latches, the operand would be available in the ARF or memory, in the converse case,  the entry in both, the ROB, and the latches, the operand is forwarded from the latches and the counter is incremented .  In the case that the entry exists in the ROB, but not in the latches, a flag would be set in the ROB to forward the result when committing to the ARF. The last case, that is the entry exists in the latches and not in the ROB, means that the result has already been committed and the result could be forwarded from the ARF.  Next in our update entries (when the cycle counter expires -- execution has completed) we would add in the new entry to the retention latch.  Lastly when the ROB entires are valid, and ready to be committed to the ARF,
if the forwarding flag is set, the forwarding counter is incremented.

\subsection{Power Tabulation Implementation}
In order for our simulation to determine the power utilization, we do not try and equate each event with an associated cost in joules.  As technology changes, and new techniques are developed to decease circuit level power usages, our simulation will become less and less accurate.  Instead we chose to simply tabulate each event that is associated with power cost, such as bit flips from 0 to 1, register comparator usage, and bit lines driven.  To help measure leakage power and overall states of the ROB, we also keep track of the time (in cycles) each bit stays in the high state, or the low state.
